Quantum Hoare Types

### 2019-10-11 paper/paper.tex
Get spacing right with code example
Hacky centering of code
Remove extra semicolon, move on to types
Add one note
Remove polymorphism to reduce complexity for first attempt at this system

\subsection{Syntax}

See Fig. \ref{fig:syntax}. Let's assume $\alpha$ represents type of a non-specific quantum state vector (that includes entangled qubits).

\begin{figure}
\begin{tabular}{lrcl}
	\textit{Types} & $A, B, C$ & ::= & $ \alpha \mid \mathrm{qbit} \mid A \otimes B \mid 1 \mid \Pi x{:}A.B \mid \Psi. X. \{P\} x{:}A \{Q\} \mid \forall \alpha.A$ \\
	\textit{Monotypes} & $\tau, \sigma$ & ::= & $ \alpha \mid \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{id}_A(M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \\
		\forall x{:}A.P \mid \forall \alpha.P \mid \forall h{:}\mathrm{heap}.P \mid \mathrm{is\_entangled}\ M $
	\end{minipage}\\
	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid K\ \tau \mid M : A $\\
	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ K \mid (\,) \mid \lambda x.\ M \mid \Lambda \alpha.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
	\end{minipage}\\
	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_A\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid \\
		\mathrm{fix}\ f(y{:}A){:}B = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
	\end{minipage}\\
	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_A M; E $\\
	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}A \mid \Delta, \alpha $\\
	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
	\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
\end{tabular}
	\caption{Syntax}
\label{fig:syntax}
\end{figure}

@@

\textit{Computations.}  All effectful stuff happens here. Separated by semicolon. Should terminate with a return value. return $M$ and $x \gets K; E$ correspond to monadic \textit{unit} and \textit{bind}.

Simple example to create a Bell Pair using QHTT commands:

\begin{tabular}{ll}
	$\mathrm{plus\_minus} =_{\mathrm{qbit} \to \mathrm{qbit}} \lambda q.\ \mathrm{do}\ ($&$a \Leftarrow \mathrm{apply}(\mathrm{Hadamard}, q);$\\
																																	&$\mathrm{return}\ a);$\\
	$\mathrm{share} =_{\mathrm{qbit} \to \mathrm{qbit} \otimes \mathrm{qbit}} \lambda a.\ \mathrm{do}\ ($&$q \Leftarrow \mathrm{alloc_{qbit}}\ \ket{0};$\\
																																							   &$b \Leftarrow \mathrm{capply}(\mathrm{a, QNot}, q);$\\
																																							   &$\mathrm{return}\ (a, b));$\\
	$a \gets \mathrm{plus\_minus} \ket{0};$ & \\
	$\mathrm{bell_{00}} \gets \mathrm{share}\ a;$ & \\
	$\mathrm{return}\ \mathrm{bell_{00}}; $&
\end{tabular}

>>>>

\subsection{Abstract Syntax}

See Fig. \ref{fig:syntax}.

\begin{figure}
\begin{tabular}{lrcl}
	\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P $
	\end{minipage}\\
	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
	\end{minipage}\\
	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
		\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
	\end{minipage}\\
	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
	\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
\end{tabular}
	\caption{Syntax}
\label{fig:syntax}
\end{figure}

@@

\textit{Commands.} \textit{apply} is for unitary circuit application (apply circuit M to N), \textit{capply} that with controls M, circuit N and arguments O. (An alternative is to formulate gate/circuit application as assignment.)

\textit{Computations.} All effectful stuff happens here. Separated by semicolon. Should terminate with a return value. return $M$ and $x \gets K; E$ correspond to monadic \textit{unit} and \textit{bind}. $x =_\tau M$ is syntactic sugar for the let-binding of $M:\tau$ to $x$.

Simple example to create a Bell Pair using QHTT commands:\\

\begin{center}
\begingroup
\setlength{\tabcolsep}{0pt}
\begin{tabular}{ll}

	$\mathrm{plus\_minus} =_{\mathrm{qbit} \to \mathrm{qbit}} \lambda q.\ \mathrm{do}\ ($&$a \Leftarrow \mathrm{apply}(\mathrm{Hadamard}, q);$\\
																																	&$\mathrm{return}\ a);$\\
\end{tabular}

\begin{tabular}{ll}
	$\mathrm{share} =_{\mathrm{qbit} \to \mathrm{qbit} \otimes \mathrm{qbit}} \lambda a.\ \mathrm{do}\ ($&$q \Leftarrow \mathrm{alloc_{qbit}}\ \ket{0};$\\
		&$b \Leftarrow \mathrm{capply}(\mathrm{a, QNot}, q);$\\
		&$\mathrm{return}\ (a, b));$\\
	$a \gets \mathrm{plus\_minus} \ket{0};$ & \\
	$\mathrm{bell_{00}} \gets \mathrm{share}\ a;$ & \\
	$\mathrm{return}\ \mathrm{bell_{00}}$&
\end{tabular}
\endgroup
\end{center}

In above, we can abbreviate the last two lines to: eval(share $a$).

\textit{Types.}
